Preregistration: Changes

### 2022-07-25 manuscript.Rmd
- language edits

As a result, with this study I want to build on this research and investigate whether or not COVID-19 related social media use meaningfully affected well-being during the pandemic.
To this end, I analyzed a large-scale **stand-alone** panel study from the Austrian Corona Panel Project [@kittelAustrianCoronaPanel2020]. 
The panel consists of **32** waves and has an overall sample size of **`r nrow(d_wide) %>% prettyNum(big.mark = ",")` participants**.
<!-- At least 1,500 participants took part per wave, and the sample is representative of the Austrian population. -->
The panel study collected a large number of psychological and demographic variables.
<!-- Being able to control for many confounding third variables, both stable and varying, together with the longitudinal design, this creates a unique opportunity  -->
I explicitly aimed to investigate the causal effects of COVID-19 related social media use on well-being.

@@

**Two underlying theories guided the selection of variables for my analysis, the two-continua model of mental health [@greenspoonIntegrationSubjectiveWellbeing2001] and the hierarchical taxonomy of computer-mediated communication [@meierComputermediatedCommunicationSocial2020a].**
According to the two-continua model, mental health consists of (a) psychopathology and (b) well-being.
<!-- Because the aim of this study is better to understand typical users and everyday contexts, my focus will be on well-being. -->
Well-being can be differentiated into subjective and psychological well-being [@dienerAdvancesOpenQuestions2018].
Whereas subjective well-being emphasizes hedonic aspects such as happiness and joy, psychological well-being addresses eudaimonic aspects such as fulfillment and meaning.
Subjective well-being is primarily about achieving positive and avoiding negative affect.
One of the most prominent indicators of well-being is life satisfaction.
In my view, because it represents a general appraisal of one's life, life satisfaction is best thought of as a meta concept combining psychological and subjective well-being.
Notably, life satisfaction is stable and fluctuates only little, whereas it's the exact opposite for affect [@dienlinImpactDigitalTechnology2020].
<!-- Everyone can feel bad on any given day, but whether or not one feels generally comfortable in one's life rather changes across years. -->
To capture well-being in this study I thus build on life satisfaction, positive affect, and negative affect.
Together, this should provide an encompassing perspective on potential media effects.

The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology.
First, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); and sixth, the message (e.g., content) [@meierComputermediatedCommunicationSocial2020a].
Whereas the first four levels focus on the communication _channel_, the last two address the communication _type_.
**Differentiating different communication channels and communication types is important, because the effects on well-being likely differ across communication channels and communication types.**
**Whereas active social media use such as chatting is routinely linked to improved well-being, passive use such as reading is often considered more negative [@dienlinImpactDigitalTechnology2020].**
**Similarly, branded apps are separate communication entities with potentially divergent effects and affordances.**
**For example, @waterlooNormsOnlineExpressions2018 found that it's more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram.**
**Especially during a pandemic, it makes sense to analyze if users engage with COVID-19 related content on Instagram, where communication is more positive, or on Facebook, where communication is more critical.**
**First studies suggest that Instagram use was indeed more beneficial for well-being than Facebook use [@masciantonioDonPutAll2021].**

**In this study, to measure the effects of social media use focused on COVID-19 related news and topics, I adopt both the channel and the type of communication perspective.**
**Together, this should offer a nuanced and comprehensive understanding of communication.**
First, I investigate how well-being is affected by different types of communication, namely active and passive use.
Defining what constitutes active and what passive use is not always clear, and different understandings are currently discussed [@ellisonWhyWeDon2020; @meierComputermediatedCommunicationSocial2020a].
Reading is generally considered as passive and writing as active, while there are also specific behaviors such as liking or sharing content that fall somewhere in-between [@meierDoesPassiveSocial2022].
In this study, I hence distinguish (a) _reading_ (passive), (b) _posting_ (active), and (c) _liking and sharing_ COVID-19 related posts (both active and passive).
Second, I analyze how using the most prominent branded applications affects well-being, and whether this effect changes across applications.
The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube---which were, at the time of writing, the most relevant social media apps.

**Worth noting, this study is not about _general_ social media use during times of COVID, but on social media use _focused_ on COVID-19 related content.** 
**For example, posting thoughts about the pandemic, reading posts and comments, or retweeting COVID-19 related news.**

@@

From a theoretical perspective, how could we explain whether COVID-19 related social media use affects well-being?
<!-- In what follows, I outline potential arguments as to why the effect might be positive or negative, direct or indirect, or nonexistent. -->
According to the set-point model of subjective well-being, well-being is surprisingly stable [@lykkenHappinessWhatStudies1999].
Although specific events such as marriage or salary can have significant effects, after some time well-being routinely returns to prior levels, which are mostly determined genetically [@sheldonStabilityHappinessTheories2014].
Only very specific events and factors such as unemployment, disability, or death can cause long-term changes in well-being [@lucasAdaptationSetpointModel2007].
<!-- So although well-being can change this does not happen easily. -->
<!-- Although the set-point theory has been criticized lately because events such as marriage, unemployment, or death can indeed lead to long-term changes, it is still widely maintained that well-being is very stable and hence hard to be influenced  -->

Can media use be such a factor?
In advance, there doesn't seem to be a clear winner, and it's likely that both positive and negative effects cancel each other out.
Empirically, social media use on average does not have a strong effect on well-being [@meierComputermediatedCommunicationSocial2020a].
According to the Differential Susceptibility to Media Effects Model [@valkenburgDifferentialSusceptibilityMedia2013], the effects of media use differ across individuals **and content**.
Whereas for some media are beneficial, for others they are harmful.
**Whereas some content provides opportunities (education, advice), other content creates risks (misinformation, hate). [@livingstoneEuropeanResearchChildren2018].**
Social media can impair well-being when causing embarrassment, stress, or disinformation, and they can improve well-being when providing connectedness, information, or entertainment [@buchiDigitalWellbeingTheory2021].
On average, however, effects are often small or negligible.

**Two of the most prominent media effect theories argue (indirectly) against strong average negative impacts.**
**According to mood management theory [@zillmannMoodManagementCommunication1988], using media substantially affects people's moods.**
**Effects can be stimulating or overwhelming, relaxing or boring.**
**After some time, users implicitly learn what media help them balance their mood and affect according to their own situational needs [@zillmannMoodManagementCommunication1988].**
**Those media that eventually become part of one's media repertoire are hence, on average, beneficial for users and their moods.**
**Using experience sampling of well-being and logs of social media use, a study with 82 participants from Italy found that after episodes of social media use, levels of positive affect increased significantly [@marcianoDynamicsAdolescentsSmartphone2022].**

**While mood management theory considers media use mainly driven by implicit learning experiences, uses and gratifications theory upholds that the process is more explicit and rational [@katzUsesGratificationsResearch1973].**
**Users select those media that they expect to have a desired effect, for example on mood, knowledge, or entertainment.** 
**If those beneficial media effects do not exist or if they are not expected, people will spend their time elsewhere.**
**And social media offer several beneficial effects, explaining why they are used that much.**
**They help find relevant information, maintain and foster relationships, express one's personality, and entertain oneself [@pelletierOneSizeDoesn2020].**
**In conclusion, because people spend so much time on social media consuming COVID-19 related content, according to both mood management theory and uses and gratifications theory we wouldn't expect to find strong average negative effects.**

But people can also misjudge media effects and are often overly optimistic [@metzgerComparativeOptimismPrivacy2017].
And precisely because social media have so many positive consequences, one can ask if this is not where the actual problem lies.
In other words, social media aren't problematic because they're inherently bad, but rather because they're too good.
And as with many other things, there can be too much of a good thing.
It is therefore often asked whether social media are addictive, and users sometimes express this fear themselves [@yangCanWatchingOnline2021]. 
However, a recently published meta-analysis found that the two most prominent measures of addiction, the Bergen Facebook Addiction Scale and the Bergen Social Media Addiction Scale, have only small relations to well-being [@duradoniWellbeingSocialMedia2020]. 
In addition, the general idea of labeling excessive social and new media use as addiction was criticized, arguing that social media represent new regular behaviors that should not be pathologized [@galerHowMuchToo2018; @vanrooijWeakScientificBasis2018].

**Because media effects can differ across users, situations, and content [@livingstoneEuropeanResearchChildren2018; @valkenburgDifferentialSusceptibilityMedia2013], I now briefly focus on the effects of COVID-19 related social media use specifically.**
First, one could assume a _direct_ negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Consuming such news can be depressing [@dornemannHowGoodBad2021], perhaps even changing some general perspectives on life.
<!-- So, just like being hit by a hammer hurts and we don't need any "mediating mechanism", this could be the case here as well. -->
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects [@dornemannHowGoodBad2021].
A study with 2.057 respondents from Italy reported that during the pandemic virtual community and social connectedness even increased during the pandemic [@guazziniSecondWaveAnalysis2022].
**In Finland, in a sample of 735 participants, levels of loneliness did not decrease during the pandemic, and people who engaged more on social media showed less loneliness [@latikkaLonelinessPsychologicalDistress2022].**
<!-- However, in light of a worldwide pandemic with millions of deaths, the negative direct effect seems more plausible. -->

There could also be _indirect_ effects.
When browsing social media for COVID-19 related news, many users reported being captivated to such an extent that they could not stop using social media [@kleinDarklySoothingCompulsion2021].
During the pandemic social media use was at an all-time high in the US [@statistaAverageDailyTime2021].
Although it is most likely that moderate social media use is not detrimental [@orbenTeenagersScreensSocial2020], overuse, however, might be more critical, and several studies have shown more pronounced negative effects for extreme users [@przybylskiLargescaleTestGoldilocks2017].
To explain, overuse could impair well-being if it replaces meaningful or functional activities such as meeting others, working, actively relaxing, or exercising.
**Another potentially negative mechanism at play are problematic social comparison processes. During the pandemic, several users shared how they successfully dealt with challenges such as physical distancing. In a study with 1131 residents from Wuhan in China [@yuePassiveSocialMedia2022], people who spent more time in quarantine also spent more time on social media. Those, who spent more time on social media also engaged in more upward social comparison, which was related to increased levels of stress.**
<!-- Therefore, if a society collectively overuses social media during a pandemic, there is potential for negative effects. -->
<!-- If overuse replaces such activities it's reasonable to assume that it's also detrimental. -->

@@

To summarize, from a theoretical perspective it is most likely that the average effects of social media use on well-being are negligible. 
Building on established theories from Communication, we would not assume that effects are either profoundly negative or strongly positive.

@@

So far, there is only little empirical research on how well-being is affected by COVID-19 related social media.
In their study on the relations between media use and mental health during the pandemic, @bendauAssociationsCOVID19Related2021 found that people who used social media as a primary source of information reported on average "significantly more unspecific anxiety and depression [] and significantly more specific COVID-19 related anxiety symptoms" (p. 288).
@edenMediaCopingCOVID192020 analyzed the media use of 425 US college students during the first wave of the pandemic, finding both positive and negative relations with well-being.
In a sample of 312 respondents collected via Amazon Mechanical Turk, @choiMediatedCommunicationMatters2021 reported that people who used media to attain information were more lonely and less satisfied with their lives.
@stainbackCOVID1924News2020 analyzed a large-scale study with 11,537 respondents from the US and found that increased COVID-19 media consumption was related to more psychological distress.
A four-wave panel study with 384 young adults from the U.S. analyzed the effects of general digital technology use---objectively measured via screenshots of screen-time applications---on mental health, separating within- and between-person relations [@sewallObjectivelyMeasuredDigital2021].
The results showed that digital technology did not have any significant effects on mental health [for a similar study with comparable results, see @bradleyStressMoodSmartphone2021].
Together, the literature is mixed, with a slight focus on the negative effects of social media as news use [see also @dornemannHowGoodBad2021; @liuRelationOfficialWhatsAppdistributed2020; @riehmAssociationsMediaExposure2020].
However, note that all of these findings represent between-person relations stemming from cross-sectional data (see below).
We therefore don't know whether the differences in mental health and well-being are due to social media use or due to other third variables, such as age, health, employment, or education.

The question of whether and how social media use affects well-being _in general_, on the other hand, is well-researched.
This also holds true for the different types of communication such as active or passive use.
A meta review (i.e., an analysis of meta-analyses) found that the relation between social media use and well-being is likely in the negative spectrum but very small [@meierComputermediatedCommunicationSocial2020a]---potentially too small to matter.
What determines whether or not an effect is considered small or trivial?
As a starting point, we could refer to standardized effect sizes.
According to @cohenPowerPrimer1992, small effect sizes start at _r_ = .10.
And indeed, several if not most of the current meta-analyses find effect sizes below that threshold [@huangTimeSpentSocial2017; @meierComputermediatedCommunicationSocial2020a; @fergusonThisMetaanalysisScreen2021].

Finally, also several individual studies employing advanced methods found smalls relations between social media use and well-being [@orbenSocialMediaEnduring2019; @przybylskiDoesTakingShort2021a; @schemerImpactInternetSocial2021; @kerestesAdolescentsOnlineSocial2020].
For example, @beyensSocialMediaUse2021 reported that although for some users (roughly one quarter) the effects of social media use on well-being were negative, for almost the same number of users they were positive, while for the rest the effects were neutral.
This finding is aligned with the Differential Susceptibility to Media Effects Model: 
Although there is substantial _variation_ of media effects for individual users, the _average_ effects reported in the literature are often small [@valkenburgDifferentialSusceptibilityMedia2013].

@@

In conclusion, in light of the theoretical considerations and empirical studies presented above, I expect that COVID-19 related communication on social media doesn't affect well-being in a meaningful or relevant way.
**This general hypothesis will be analyzed specifically for the communication types of (a) time spent reading, (b) liking and sharing, and (c) actively posting COVID-19 related content.**
**In addition, I will analyze how well-being is influenced by spending time on five prominent social media apps, including (a) Facebook, (b) Instagram, (c) Twitter, (d) WhatsApp, and (e) YouTube.** 
**Three different well-being indicators will be differentiated: life satisfaction, positive affect, and negative affect.**

@@

@@

What does this mean practically and how can it be operationalized?
In this study, COVID-19 related social media use was measured on a 5-point scale, ranging from 1 = _never_ to 5 = _several times a day_. 
Thus, a change of four units in social media use (e.g., a complete stop) should correspond to a noticeable change in well-being.
But what's a noticeable change in well-being?
According to @normanInterpretationChangesHealthrelated2003, people can reliably distinguish _seven_ levels of satisfaction with health.
So if satisfaction is measured on a 7-point scale, we would state that a four unit change in social media use should result in a one unit change in life satisfaction. 
(For more information, see Methods section "Inference Criteria.")

@@

The hypothesis explicitly states a causal effect.
In non-experimental studies, longitudinal designs can help investigate causality.
Using longitudinal designs alone, however, is not sufficient for establishing correct causal statements [@rohrerTheseAreNot2021].
In addition, we for example also need to control for confounding third variables.
**Importantly, when analyzing longitudinal (within-person) relationships and effects, it is important to control for _varying_ third variables.**
**Non-varying third variables only help control non-varying (between-person) relations.**

@@

When searching for suitable candidates for confounders, we should look for variables that affect both media use and well-being. 
Controlling for these factors isolates the actual effect of social media use on well-being.
We can also control for variables that affect only social media use or well-being. 
However, in doing so not much is gained or lost, because the effects of social media use would remain virtually the same [@klinePrinciplesPracticeStructural2016; but see @mcelreathYesterdayClass2021].

@@

Next to including covariates, it's now increasingly understood that causal effects should be analyzed from an internal, within-person perspective [@hamakerWhyResearchersShould2014].
If a specific person changes their media diet, we need to measure how this behavior affects _their own_ well-being.
Between-person comparisons from cross-sectional data, where participants are interviewed only once, cannot provide such insights.
In this study, I hence differentiate between-person relations from within-person effects. 
And as explicated above, to test the hypothesis I thus consider only the within-person effects.

>>>>

As a result, with this study I want to build on this research and investigate whether or not COVID-19 related social media use meaningfully affected well-being during the pandemic.
To this end, I analyzed a large-scale **standalone** panel study from the Austrian Corona Panel Project [@kittelAustrianCoronaPanel2020]. 
The panel consists of **32** waves and has an overall sample size of **`r nrow(d_wide) %>% prettyNum(big.mark = ",")` participants**.
<!-- At least 1,500 participants took part per wave, and the sample is representative of the Austrian population. -->
The panel study collected a large number of psychological and demographic variables.
<!-- Being able to control for many confounding third variables, both stable and varying, together with the longitudinal design, this creates a unique opportunity  -->
I explicitly aimed to investigate the causal effects of COVID-19 related social media use on well-being.

@@

**Two underlying theories guided the selection of variables for this study, namely the two-continua model of mental health [@greenspoonIntegrationSubjectiveWellbeing2001] and the hierarchical taxonomy of computer-mediated communication [@meierComputermediatedCommunicationSocial2020a].**
According to the two-continua model, mental health consists of (a) psychopathology and (b) well-being.
<!-- Because the aim of this study is better to understand typical users and everyday contexts, my focus will be on well-being. -->
Well-being can be differentiated into subjective and psychological well-being [@dienerAdvancesOpenQuestions2018].
Whereas subjective well-being emphasizes hedonic aspects such as happiness and joy, psychological well-being addresses eudaimonic aspects such as fulfillment and meaning.
Subjective well-being is primarily about achieving positive and avoiding negative affect.
One of the most prominent indicators of well-being is life satisfaction.
In my view, because it represents a general appraisal of one's life, life satisfaction is best thought of as a meta concept combining psychological and subjective well-being.
Notably, life satisfaction is stable and fluctuates only little, whereas it's the exact opposite for affect [@dienlinImpactDigitalTechnology2020].
<!-- Everyone can feel bad on any given day, but whether or not one feels generally comfortable in one's life rather changes across years. -->
To capture well-being in this study I thus build on life satisfaction, positive affect, and negative affect.
Together, this should provide an encompassing perspective on potential media effects.

The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology.
First, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); and sixth, the message (e.g., content) [@meierComputermediatedCommunicationSocial2020a].
Whereas the first four levels focus on the communication _channel_, the last two address the communication _type_.
**Distinguishing different communication channels and communication types is important, because the effects on well-being likely differ across communication channels and communication types.**
**Whereas active social media use such as chatting is routinely linked to improved well-being, passive use such as reading is often considered more negative [@dienlinImpactDigitalTechnology2020].**
**Similarly, branded apps are separate communication entities with potentially divergent effects and affordances.**
**For example, @waterlooNormsOnlineExpressions2018 found that it's more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram.**
**Especially during a pandemic, it makes sense to analyze if users engage with COVID-19 related content on Instagram, where communication is more positive, or on Facebook, where communication is more critical.**
**First studies suggest that during the pandemic Instagram use was indeed more beneficial for well-being than Facebook use [@masciantonioDonPutAll2021].**
**In this study, to measure the effects of social media use focused on COVID-19 related news and topics, I adopt both the channel and the type of communication perspective.**
<!-- **Together, this should offer a nuanced and comprehensive understanding of communication.** -->

Specifically, I investigate how well-being is affected by different types of communication, namely active and passive use.
Defining what constitutes active and what passive use is not always clear, and different understandings are currently discussed [@ellisonWhyWeDon2020; @meierComputermediatedCommunicationSocial2020a].
Reading is generally considered as passive and writing as active use, while there are also specific behaviors falling somewhere in-between such as liking or sharing content [@meierDoesPassiveSocial2022].
**In this study, I hence distinguish (a) reading (passive), (b) posting (active), and (c) liking and sharing COVID-19 related posts (both active and passive).**
**Second, I analyze how using the most prominent branded applications affects well-being, and whether this effect changes across applications.**
**The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube---which were, at the time of writing, the most relevant social media apps.**

**Worth noting, this study is not about _general_ social media use during times of COVID, but on social media use _focused_ on COVID-19 related content.** 
**This, for example, includes posting thoughts about the pandemic, reading posts and comments, or retweeting COVID-19 related news.**

@@

From a theoretical perspective, how could we explain whether and how the various types of COVID-19 related social media use affects well-being?
<!-- In what follows, I outline potential arguments as to why the effect might be positive or negative, direct or indirect, or nonexistent. -->
According to the set-point model of subjective well-being, well-being is surprisingly stable [@lykkenHappinessWhatStudies1999].
Although specific events such as marriage or salary can have significant effects, after some time well-being routinely returns to prior levels, which are mostly determined genetically [@sheldonStabilityHappinessTheories2014].
Only very specific events and factors such as unemployment, disability, or death can cause long-term changes in well-being [@lucasAdaptationSetpointModel2007].
<!-- So although well-being can change this does not happen easily. -->
<!-- Although the set-point theory has been criticized lately because events such as marriage, unemployment, or death can indeed lead to long-term changes, it is still widely maintained that well-being is very stable and hence hard to be influenced  -->

Can media use be such a factor?
In advance, there doesn't seem to be a clear winner, and it's likely that both positive and negative effects cancel each other out.
Empirically, social media use on average does not have a strong effect on well-being [@meierComputermediatedCommunicationSocial2020a].
According to the Differential Susceptibility to Media Effects Model [@valkenburgDifferentialSusceptibilityMedia2013], the effects of media use differ across individuals **and types of content**.
Whereas for some media are beneficial, for others they are harmful.
**Whereas some content mostly provides opportunities (education, advice), other content rather creates risks (misinformation, hate) [@livingstoneEuropeanResearchChildren2018].**
Social media can impair well-being when causing embarrassment, stress, or disinformation, and they can improve well-being when providing connectedness, information, or entertainment [@buchiDigitalWellbeingTheory2021].
On average, however, effects are often small or negligible.

**Two prominent media effect theories argue (mostly implicitly) against strong average negative effects.**
**According to mood management theory [@zillmannMoodManagementCommunication1988], using media can substantially affect people's moods.**
**Use can be stimulating or overwhelming, relaxing or boring.**
**After some time, users implicitly learn what media help them balance their mood and affect according to their own situational needs [@zillmannMoodManagementCommunication1988].**
**Those media that eventually become part of one's media repertoire are hence, on average, beneficial for users to regulate their mood.**
**Using experience sampling of well-being and logs of social media use, a study with 82 participants from Italy found that after episodes of social media use, levels of positive affect increased significantly [@marcianoDynamicsAdolescentsSmartphone2022].**

**While mood management theory considers media use mainly driven by implicit learning experiences, uses and gratifications theory upholds that the process is more explicit and rational [@katzUsesGratificationsResearch1973].**
**Users select those media that they expect to have a desired effect, for example on mood, knowledge, or entertainment.** 
**If those beneficial media effects do not exist or if they are not expected, people will spend their time elsewhere.**
**And social media offer several beneficial effects, explaining why they are used that much.**
**They help find relevant information, maintain and foster relationships, express one's personality, and entertain oneself [@pelletierOneSizeDoesn2020].**
**In conclusion, because people spend so much time on social media consuming COVID-19 related content, according to both mood management theory and uses and gratifications theory this is indirect proof that average effects on well-being are likely not particularly negative.**

But people can also misjudge media effects and are often overly optimistic [@metzgerComparativeOptimismPrivacy2017].
Precisely because social media have so many positive consequences, one can ask if this is not where the actual problem lies.
In other words, social media might not problematic because they are inherently bad, but rather because they are too good.
As with many other things, there can be too much of a good thing.
It is therefore often asked whether social media can become addictive, and users sometimes express this fear themselves [@yangCanWatchingOnline2021]. 
However, a recently published meta-analysis found that the two most prominent measures of addiction, the Bergen Facebook Addiction Scale and the Bergen Social Media Addiction Scale, have only small relations to well-being [@duradoniWellbeingSocialMedia2020]. 
In addition, the general idea of labeling excessive social and new media use as addiction was criticized, arguing that social media represent new regular behaviors that should not be pathologized [@galerHowMuchToo2018; @vanrooijWeakScientificBasis2018].

**Because media effects can differ across users, situations, and content [@livingstoneEuropeanResearchChildren2018; @valkenburgDifferentialSusceptibilityMedia2013], I now briefly focus on the effects of COVID-19 related social media use specifically.**
First, one could assume a _direct_ negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Consuming such news can be depressing [@dornemannHowGoodBad2021], perhaps even changing some general perspectives on life.
<!-- So, just like being hit by a hammer hurts and we don't need any "mediating mechanism", this could be the case here as well. -->
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects [@dornemannHowGoodBad2021].
A study with 2.057 respondents from Italy reported that during the pandemic virtual community and social connectedness even increased [@guazziniSecondWaveAnalysis2022].
**In a study with 735 participants from Finland, levels of loneliness did not decrease during the pandemic, and people who engaged more on social media experienced less loneliness [@latikkaLonelinessPsychologicalDistress2022].**
<!-- However, in light of a worldwide pandemic with millions of deaths, the negative direct effect seems more plausible. -->

Second, there could also be _indirect_ effects.
When browsing social media for COVID-19 related news, many users reported being captivated to such an extent that they could not stop using social media [@kleinDarklySoothingCompulsion2021].
During the pandemic social media use was at an all-time high in the US [@statistaAverageDailyTime2021].
Although it is most likely that moderate social media use is not detrimental [@orbenTeenagersScreensSocial2020], overuse, however, might be more critical, and several studies have shown more pronounced negative effects for extreme users [@przybylskiLargescaleTestGoldilocks2017].
To explain, overuse could impair well-being if it replaces meaningful or functional activities such as meeting others, working, actively relaxing, or exercising.
**Another potentially negative mechanism at play are problematic social comparison processes. During the pandemic, several users shared how they successfully dealt with challenges such as physical distancing. In a study with 1,131 residents from Wuhan in China [@yuePassiveSocialMedia2022], people who spent more time in quarantine also spent more time on social media. Those, who spent more time on social media also engaged in more upward social comparison, which was related to increased levels of stress.**
<!-- Therefore, if a society collectively overuses social media during a pandemic, there is potential for negative effects. -->
<!-- If overuse replaces such activities it's reasonable to assume that it's also detrimental. -->

@@

To summarize, it seems that from a theoretical perspective it is most likely that the average effects of social media use on well-being are negligible. 
Building on established theories from Communication and current empirical findings, we would not assume that effects are either profoundly negative or strongly positive.

@@

So far, there is still comparatively little empirical research on how well-being is affected by COVID-19 related social media use.
In their study on the relations between media use and mental health during the pandemic, @bendauAssociationsCOVID19Related2021 found that people who used social media as a primary source of information reported on average "significantly more unspecific anxiety and depression [] and significantly more specific COVID-19 related anxiety symptoms" (p. 288).
@edenMediaCopingCOVID192020 analyzed the media use of 425 US college students during the first wave of the pandemic, finding both positive and negative relations with well-being.
In a sample of 312 respondents collected via Amazon Mechanical Turk, @choiMediatedCommunicationMatters2021 reported that people who used media to attain information were more lonely and less satisfied with their lives.
@stainbackCOVID1924News2020 analyzed a large-scale study with 11,537 respondents from the US and found that increased COVID-19 media consumption was related to more psychological distress.
A four-wave panel study with 384 young adults from the U.S. analyzed the effects of general digital technology use---objectively measured via screenshots of screen-time applications---on mental health, separating within- and between-person relations [@sewallObjectivelyMeasuredDigital2021].
The results showed that digital technology did not have significant effects on mental health [for a similar study with comparable results, see @bradleyStressMoodSmartphone2021].
Together, the literature is mixed, with a slight focus on the negative effects of social media as news use [see also @dornemannHowGoodBad2021; @liuRelationOfficialWhatsAppdistributed2020; @riehmAssociationsMediaExposure2020].
<!-- However, note that all of these findings represent between-person relations stemming from cross-sectional data (see below). -->
<!-- We therefore don't know whether the differences in mental health and well-being are due to social media use or due to other third variables, such as age, health, employment, or education. -->

The question of whether and how social media use affects well-being _in general_, on the other hand, is well-researched.
This also holds true for the different types of communication such as active or passive use.
A meta review (i.e., an analysis of meta-analyses) found that the relation between social media use and well-being is likely in the negative spectrum but very small, potentially too small to matter [@meierComputermediatedCommunicationSocial2020a].
What determines whether or not an effect should be considered small or trivial?
As a starting point, we could refer to standardized effect sizes.
According to @cohenPowerPrimer1992, small effect sizes start at _r_ = .10.
And indeed, several if not most of the current meta-analyses find effect sizes below that threshold [@huangTimeSpentSocial2017; @meierComputermediatedCommunicationSocial2020a; @fergusonThisMetaanalysisScreen2021].

Finally, several individual studies employing advanced methods found smalls relations between social media use and well-being [@orbenSocialMediaEnduring2019; @przybylskiDoesTakingShort2021a; @schemerImpactInternetSocial2021; @kerestesAdolescentsOnlineSocial2020].
For example, @beyensSocialMediaUse2021 reported that although for roughly one quarter of all users the effects of social media use on well-being were negative, for almost the same number of users they were positive, while for the rest the effects were neutral.
This finding is aligned with the Differential Susceptibility to Media Effects Model: 
Although there is substantial _variation_ of media effects for individual users, the _average_ effects reported in the literature are often small [@valkenburgDifferentialSusceptibilityMedia2013].

@@

In conclusion, in light of the theoretical considerations and empirical studies presented above, I expect that COVID-19 related communication on social media doesn't affect well-being in a meaningful or relevant way.

@@

**In this study, this general hypothesis will be analyzed specifically for the three communication types of (a) time spent reading, (b) liking and sharing, and (c) actively posting COVID-19 related content.**
**In addition, I will analyze how well-being is influenced by spending time on five prominent social media apps, including (a) Facebook, (b) Instagram, (c) Twitter, (d) WhatsApp, and (e) YouTube.** 
**Three different well-being indicators will be distinguished: life satisfaction, positive affect, and negative affect.**

@@

What does this mean practically and how can it be operationalized?
In this study, COVID-19 related social media use was measured on a 5-point scale, ranging from 1 = _never_ to 5 = _several times a day_. 
Thus, a change of four units in social media use (e.g., a complete stop) should correspond to a noticeable change in well-being.
But what's a noticeable change in well-being?
According to @normanInterpretationChangesHealthrelated2003, people can reliably distinguish seven levels of satisfaction with health.
So if satisfaction is measured on a 7-point scale, we would state that a four unit change in social media use should result in a one unit change in life satisfaction. 
(For more information, see Methods section "Inference Criteria.")

@@

The hypothesis explicitly states a causal effect.
In non-experimental studies, longitudinal designs can help investigate causality.
Using longitudinal designs alone, however, is not sufficient for establishing correct causal statements [@rohrerTheseAreNot2021].
In addition, we for example also need to control for confounding third variables.
**Importantly, when analyzing longitudinal (within-person) relationships and effects, it is important to control for _varying_ third variables.**
**Non-varying third variables can only help control non-varying (between-person) relations.**

@@

When searching for suitable candidates for confounders, we should look for variables that affect both media use and well-being. 
Controlling for these factors isolates the actual effect of social media use on well-being.
We can also control for variables that affect only social media use or well-being. 
However, in doing so not much is gained, because the effects of social media use would remain virtually the same [@klinePrinciplesPracticeStructural2016; but see @mcelreathYesterdayClass2021].

@@

Next to including covariates, it's now increasingly understood that causal effects should be analyzed from an internal, within-person perspective [@hamakerWhyResearchersShould2014].
If a specific person changes their media diet, we need to measure how this behavior affects their own well-being.
Between-person comparisons from cross-sectional data, where participants are interviewed only once, cannot provide such insights.
In this study, I hence differentiate between-person relations from within-person effects. 
And as explicated above, to test the hypothesis I thus consider only the within-person effects.

Finally, one precondition of causality is temporal order. 
The cause needs to precede the effect.
Finding the right interval between cause and effect is crucial.
For example, if we want to understand the effect of alcohol consumption on driving performance, it makes a big difference if driving performance is measured one minute, one hour, one day, or one week after consumption.
<!-- Finding the right interval is difficult. -->
**If variables are stable, longer intervals are needed; if they fluctuate, shorter intervals.**
**In the case of well-being, we need shorter intervals for the more fluctuating positive and negative affect, and longer ones for the more stable life satisfaction [@dienlinImpactDigitalTechnology2020].**
**Using social media can have instant effects on mood [@marcianoDynamicsAdolescentsSmartphone2022].**
**Effects on life satisfaction often take longer to manifest, for example because media use leads to actual changes in specific behaviors, which then in turn affect life satisfaction [@dienlinDisplacementReinforcementReciprocity2017].**
Choosing the right interval is challenging, because especially short intervals are hard to implement in practice, often requiring advanced methods such as experience sampling (also known as in situ measurement or ambulant assessment) [@schnauber-stockmannMobileDevicesTools2020].
In this study, I hence analyze how using social media during the last week affected positive and negative affect during the same week.
In other words, if people during the last week engaged in more COVID-19 related social media use than they usually do, did they feel better or worse during that week than they usually do?
Regarding life satisfaction, I implemented a longer interval.
If people during the last week used COVID-19 related social media more than they usually do, were they at the end of the week more or less satisfied with their lives than they usually are?
I hence analyze if when a person changes their social media diet, are there (a) _simultaneous_ changes in their affect and (b) _subsequent_ changes in their life satisfaction?
**These relations will be controlled for varying confounders, which fosters a causal interpretation.**
Similar approaches were implemented by other studies [@johannesNoEffectDifferent2022; @scharkowHowSocialNetwork2020], and they are considered a best practice approach toward analyzing causality [@bellFixedRandomEffects2019].
